A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-02-20 tex/body.tex
otherk ind of l
more

%\ourparagraph{Specifications.}
In an SMT problem, the signature is typically given by specifying first the
uninterpreted ordinary types in any order, then the (co)datatypes with their constructors
and selectors in groups of mutually (co)recursive groups of (co)datatypes, and
finally any other function symbols.
%
Each (co)datatype specification consists of $l$~mutually recursive types that are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
We allow ourselves some notational parameterization
% at the metalevel
through subscripts%
\begin{conf}
(e.g., $\ty{stream}_{\,\tau}$)%
\end{conf}%
\begin{rep}---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}\end{rep}.

\newcommand\elll{\kern.18ex l\kern.11ex}
\newcommand\elllx{\kern.11ex l\kern.18ex}

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by CVC4 and Z3, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}= \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}})} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,n_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.
%
For types with several constructors, it is customary to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. Instead of extending $\Funcs$,
we let $\const{d}_{i\negvthinspace j}(t)$
abbreviate
$t \teq \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(t), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(t)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

@@

This work is part of a program that aims at enriching automatic provers
with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, it would be useful to implement
proof reconstruction for (co)data\-type inferences in Isabelle.
CVC4 cound be interfaced for generating counterexamples in proof assistants, as
an alternative to Nitpick; in this context, the acyclicity and uniqueness
rules are crucial to exclude spurious models. Finally, it might be
worthwhile to extend SMT solvers with dedicated support for (co)recursive
functions.

>>>>

%\ourparagraph{Specifications.}
In an SMT problem, the signature is typically given by specifying first the
uninterpreted ordinary types in any order, then the (co)datatypes with their constructors
and selectors in groups of mutually (co)recursive groups of (co)datatypes, and
finally any other function symbols.
%
Each (co)datatype specification consists of $\ell$~mutually recursive types that are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
We allow ourselves some notational parameterization
% at the metalevel
through subscripts%
\begin{conf}
(e.g., $\ty{stream}_{\,\tau}$)%
\end{conf}%
\begin{rep}---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}\end{rep}.

\newcommand\elll{\kern.05ex \ell\kern.05ex}
\newcommand\elllx{\kern.05ex \ell\kern.05ex}

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by CVC4 and Z3, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}})} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,n_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.
%
For types with several constructors, it is customary to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. Instead of extending $\Funcs$,
we let $\const{d}_{i\negvthinspace j}(t)$
abbreviate
$t \teq \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(t), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(t)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

@@

This work is part of a program that aims at enriching automatic provers
with high-level features and at reducing the gap between automatic and
interactive theorem proving. As future work, it would be useful to implement
proof reconstruction for (co)data\-type inferences in Isabelle.
CVC4 cound be interfaced for generating counterexamples in proof assistants, as
an alternative to Nitpick; in this context, the acyclicity and uniqueness
rules are crucial to exclude spurious countermodels. Finally, it might be
worthwhile to extend SMT solvers with dedicated support for (co)recursive
functions.
